---
title: Vignette enrigo package
author: Detlef Groth, University of Potsdam, Germany
date: 2023-09-03 09:40
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Vignette enrigo package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
include-before: |
    <style> 
    body { max-width: 1000px; font-family: Candara, sans-serif;} 
    table { min-width: 400px; border-spacing: 5px;  border-collapse: collapse; }
    </style>
---

```{r echo=FALSE,results="hide"}
library(knitr)
```

# Introduction

This is tutorial for the enrigo package which allow the  enrichment  analysis
of gene  ontology  terms. The  package  contains  two R objects,  the `goutil`
object which allows you to work with the information provided in gene ontology
files in the Obo file format and the `enr` object  which uses the  information
in so called gene ontology association files (gaf), files containing a mapping
between gene identifiers and GO identifiers.

# goutil object

## Initialization

Let's  first load the  package  and have an overview  about the methods of the
`goutil`  object.  Before you can start using the methods you have to download
gene ontology Obo       files       from       the       website
[https://www.gene-ontology.org](https://www.gene-ontology.org).

There is the method  `goutil$download`  which  helps with this. You can either
give a year or a concrete  version string like  `2023-01-01`. If you just give
the year the  first  version  released  in that  year  will be  download.  The
function  returns the file name which was  downloaded.  Let's download the OBo
files from January 2020 and 2023 for instance.


```{r}
library(enrigo)
getwd()
list.files()
list.files("..")
goutil$download(2020,quiet=TRUE)
goutil$download(2023,quiet=TRUE)
list.files()
```

After your  downloaded  the file you can initialize  your `goutil` object with
this new file. If you do this the first  time R will  parse that file and will
extract  the  required  time. This  might  take a few  minutes as R is not the
fastest  text  parser. In  subsequent  analysis  this will be much faster as R
caches the parsing result in an RDS file in parallel to the Obo file. 

```{r}
### measure the processing time
t1=Sys.time()
goutil$new("2023-01-01-go.obo")
print(Sys.time()-t1)
list.files()
```

As you can see in  addition to the obo file there is now a RDS file which will
be used in subsequence initializations, thus saving processing time.

```{r}
t1=Sys.time()
goutil$new("2023-01-01-go.obo")
print(Sys.time()-t1)
```

After  initialization you can start using the methods provided by the `goutil`
object. Let's first get a list of the existing functions.

```{r}
ls(goutil)
```

Basically  you can  distinguish  between 3 different  type of functions in the
`goutil` object.

* initialization functions, download, new, read.obofile
* retrieving information about GO ids, getName, getNamespace, getTree etc
* plotting or image functions, kroki, quickGO
* obo file functions, getOboVersion, getStats

Let's continue with the information functions

## Entry information

You can extract  information  about  individual  GO id(s) using  methods  like
getEntry, getName, getNamespace and others. Here some examples:

```{r}
childs=goutil$getChildren('GO:0008150')
for (child in childs) {
    cat(child,"\t",goutil$getNamespace(child),"\t",goutil$getName(child),"\n",sep="")
}
```

Individual  entries  from the text base can be  retrieved  using the  getEntry
function.

```{r}
cat(goutil$getEntry("GO:0000001"))
```

## Graph information

Functions like getParent, getChildren, isChild, isParent, getTree, getTreeMatrix
can be used to access  information  about the  position of a GO id in the gene
ontology graph. Here an example:

```{r}
goutil$getName('GO:0000001')
goutil$getParent('GO:0000001')
goutil$getChildren('GO:0000001')
goutil$isChild('GO:0000001','GO:0000002')
```

To get all parent  nodes for a given gene  ontology  id we can use the getTree
method  which returns all parent nodes and their parents recursively up to the
root level of the hierarchy.

```{r}
goutil$getTree('GO:0000001')
```

This just list all of these GO ids, to get an  adjacency  matrix for the graph
structure you can use the function getTreeMatrix:

```{r}
options(width=90)
goutil$getTreeMatrix('GO:0000001')
```

This might be helpful in case you like to visualize the data using a R package
like  `igraph`.  There are  however  as well two plot  function  in the `goutil`
object  available  which are using two  webservices which are described in the
next section.

## Plotting and image functions

There are two functions,  `goutil$quickGO` using the EBI QuickGO webservice and
`goutil$kroki` using the kroki webservice to visualize a Gene Ontology graph.

They can be used to retrieve the URL for the given Gene Ontology identifier or
to plot  directly  onto a R plot, the latter  required  the R png  package to be
installed on the machine, so before you can do this you should do a:

```{r eval=FALSE}
install.package("png")
```

Here an example to retrieve the URL's for the the GO identifier GO:0003676:

```{r}
goutil$quickGO('GO:0003676')
goutil$kroki('GO:0003676')
```

These  URL's  can be embedded  in your  Markdown  or HTML  document  for
instance like this (Markdown syntax for the QuickGO service):

```
### remove space before r
![](` r goutil$quickGO('GO:0003676')`)
```

A GO graph image retrieved from the Kroki  webservice using PlantUML syntax in
the background:

![](`r goutil$kroki('GO:0003676')`)



```{r fig=TRUE,fig.width=6,fig.height=6}
goutil$quickGO('GO:0003676',plot=TRUE)
```

```{r fig=TRUE,fig.width=6,fig.height=6}
goutil$kroki('GO:0003676',plot=TRUE)
```

## Gene Ontology statistics

There are two functions  which can be used to retrieve  information  about the
currently used GO file:

```{r}
goutil$getOboVersion()
goutil$getStats()
```

The same can be embedded as well directly into the text using for instance the
`knitr::kable` function:


```{r results="asis"}
knitr::kable(goutil$getStats(),
    caption=paste("**Statistics for GO version ",goutil$getOboVersion(),"**",sep=""))
```

# enr object

Now a short review of the functions available by the `enr` object.

There is as well here a download  function  which  allows you to download  the
Gene Ontology annotation files (GAF) either from the GO website - 
[https://geneontology.org/docs/download-go-annotations/](https://geneontology.org/docs/download-go-annotations/)
or  the GAF files  from         the         Hymenoptera         Genome         Database
[https://hymenoptera.elsiklab.missouri.edu/](https://hymenoptera.elsiklab.missouri.edu/),
which as well provides GAF files here: [https://data.elsiklab.missouri.edu/downloads/hgd/HGD-GO-Annotation/gaf/](https://data.elsiklab.missouri.edu/downloads/hgd/HGD-GO-Annotation/gaf/)

To see the  available  species you can just call the function `gaf` of the enr
object.

```{r}
sort(names(enr$gaf()))
```


