---
title: Vignette enrigo package
author: Detlef Groth, University of Potsdam, Germany
date: 2023-09-05 06:22
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Vignette enrigo package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
include-before: |
    <style> 
    body { max-width: 1000px; font-family: Candara, sans-serif;} 
    table { min-width: 400px; border-spacing: 5px;  border-collapse: collapse; }
    </style>
---

```{r echo=FALSE,results="hide"}
library(knitr)
```

# Introduction

This is tutorial for the enrigo package which allow the  enrichment  analysis
of gene  ontology  terms. The  package  contains  two R objects,  the `goutil`
object which allows you to work with the information provided in gene ontology
files in the Obo file format and the `enr` object  which uses the  information
in so called gene ontology association files (gaf), files containing a mapping
between gene identifiers and GO identifiers.

If you are only interested how to actually perform the enrichment analysis you
should jump to the section [enrichment](#enrichment).

# goutil object

## Initialization

Let's  first load the  package  and have an overview  about the methods of the
`goutil`  object.  Before you can start using the methods you have to download
gene ontology Obo       files       from       the       website
[https://www.gene-ontology.org](https://www.gene-ontology.org).

There is the method  `goutil$download`  which  helps with this. You can either
give a year or a concrete  version string like  `2023-01-01`. If you just give
the year the  first  version  released  in that  year  will be  download.  The
function  returns the file name which was  downloaded.  Let's download the OBo
files from January 2020 and 2023 for instance.


```{r}
library(enrigo)
getwd()
list.files()
list.files("..")
goutil$download(2020,quiet=TRUE)
goutil$download(2023,quiet=TRUE)
list.files()
```

After your  downloaded  the file you can initialize  your `goutil` object with
this new file. If you do this the first  time R will  parse that file and will
extract  the  required  time. This  might  take a few  minutes as R is not the
fastest  text  parser. In  subsequent  analysis  this will be much faster as R
caches the parsing result in an RDS file in parallel to the Obo file. 

```{r}
### measure the processing time
t1=Sys.time()
goutil$new("2023-01-01-go.obo")
print(Sys.time()-t1)
list.files()
```

As you can see in  addition to the obo file there is now a RDS file which will
be used in subsequence initializations, thus saving processing time.

```{r}
t1=Sys.time()
goutil$new("2023-01-01-go.obo")
print(Sys.time()-t1)
```

After  initialization you can start using the methods provided by the `goutil`
object. Let's first get a list of the existing functions.

```{r}
ls(goutil)
```

Basically  you can  distinguish  between 3 different  type of functions in the
`goutil` object.

* initialization functions, download, new, read.obofile
* retrieving information about GO ids, getName, getNamespace, getTree etc
* plotting or image functions, kroki, quickGO
* obo file functions, getOboVersion, getStats

Let's continue with the information functions

## Entry information

You can extract  information  about  individual  GO id(s) using  methods  like
getEntry, getName, getNamespace and others. Here some examples:

```{r}
childs=goutil$getChildren('GO:0008150')
for (child in childs) {
    cat(child,"\t",goutil$getNamespace(child),"\t",goutil$getName(child),"\n",sep="")
}
```

Individual  entries  from the text base can be  retrieved  using the  getEntry
function.

```{r}
cat(goutil$getEntry("GO:0000001"))
```

## Graph information

Functions like getParent, getChildren, isChild, isParent, getTree, getTreeMatrix
can be used to access  information  about the  position of a GO id in the gene
ontology graph. Here an example:

```{r}
goutil$getName('GO:0000001')
goutil$getParent('GO:0000001')
goutil$getChildren('GO:0000001')
goutil$isChild('GO:0000001','GO:0000002')
```

To get all parent  nodes for a given gene  ontology  id we can use the getTree
method  which returns all parent nodes and their parents recursively up to the
root level of the hierarchy.

```{r}
goutil$getTree('GO:0000001')
```

This just list all of these GO ids, to get an  adjacency  matrix for the graph
structure you can use the function getTreeMatrix:

```{r}
options(width=90)
goutil$getTreeMatrix('GO:0000001')
```

This might be helpful in case you like to visualize the data using a R package
like  `igraph`.  There are  however  as well two plot  function  in the `goutil`
object  available  which are using two  webservices which are described in the
next section.

## Plotting and image functions

There are two functions,  `goutil$quickGO` using the EBI QuickGO webservice and
`goutil$kroki` using the kroki webservice to visualize a Gene Ontology graph.

They can be used to retrieve the URL for the given Gene Ontology identifier or
to plot  directly  onto a R plot, the latter  required  the R png  package to be
installed on the machine, so before you can do this you should do a:

```{r eval=FALSE}
install.package("png")
```

Here an example to retrieve the URL's for the the GO identifier GO:0003676:

```{r}
goutil$quickGO('GO:0003676')
goutil$kroki('GO:0003676')
```

These  URL's  can be embedded  in your  Markdown  or HTML  document  for
instance like this (Markdown syntax for the QuickGO service):

```
### remove space before r
![](` r goutil$quickGO('GO:0003676')`)
```

A GO graph image retrieved from the Kroki  webservice using PlantUML syntax in
the background:

![](`r goutil$kroki('GO:0003676')`)



```{r fig=TRUE,fig.width=6,fig.height=6}
goutil$quickGO('GO:0003676',plot=TRUE)
```

```{r fig=TRUE,fig.width=6,fig.height=6}
goutil$kroki('GO:0003676',plot=TRUE)
```

## Gene Ontology statistics

There are two functions  which can be used to retrieve  information  about the
currently used GO file:

```{r}
goutil$getOboVersion()
goutil$getStats()
```

The same can be embedded as well directly into the text using for instance the
`knitr::kable` function:


```{r results="asis"}
knitr::kable(goutil$getStats(),
    caption=paste("**Statistics for GO version ",goutil$getOboVersion(),"**",sep=""))
```

# enr object

Now a short review of the functions available by the `enr` object.

## Initialization

There is as well here a download  function  which  allows you to download  the
Gene Ontology annotation files (GAF) either from the GO website - 
[https://geneontology.org/docs/download-go-annotations/](https://geneontology.org/docs/download-go-annotations/)
or  the GAF files  from         the         Hymenoptera         Genome         Database
[https://hymenoptera.elsiklab.missouri.edu/](https://hymenoptera.elsiklab.missouri.edu/),
which as well provides GAF files here: [https://data.elsiklab.missouri.edu/downloads/hgd/HGD-GO-Annotation/gaf/](https://data.elsiklab.missouri.edu/downloads/hgd/HGD-GO-Annotation/gaf/)

To see the  available  species you can just call the function `gaf` of the enr
object.

```{r}
sort(names(enr$gaf()))
```


We can as well display all species and filenames in a table:

```{r results="asis"}
tab=do.call(rbind,enr$gaf())
colnames(tab)[1]="URL"
knitr::kable(tab)
```

<a name="enrichment"> </a>

## Enrichment analysis

Let's now perform an enrichment  analysis for GO terms. Usually we have a list
of genes and some of these genes might be  significantly  changed between some
conditions.  Often we would like to group these  changed,  often  up-regulated
genes and for this  purpose GO terms can be used. We are  interested  which GO
terms  appear in this list of  (possibly  up-regulated)  genes more often than
expected by chance, i.e. they are enriched. 



So for the enrichment analysis we need a list of genes, one is the list of all
genes we were  evaluating,  the  other is a subset  of this  list, the list of
genes  which is  changed  under a certain  conditions,  as I said of this list
considers up- or  down-regulated  genes. The second thing we need is a mapping
of genes to GO terms. Such mappings are provided via GO annotation files. This
package supports annotation files from the Gene Ontology website, 
[https://geneontology.org/docs/download-go-annotations/](https://geneontology.org/docs/download-go-annotations/)
or from the Hymenoptera Genome Database
[https://hymenoptera.elsiklab.missouri.edu/](https://hymenoptera.elsiklab.missouri.edu/).
Beside of this we need a GO obo file  where the  release  file is close to the
release date of the annotation file. Let's start first with downloading the GO
obo file and the GO annotation  file. As an example we will annotate  genes of
*Apis mellifera*.


```{r}
### download and initialize goutil object with GO obofile from January 2022
fname=goutil$download(2022)
goutil$new(fname)
### download and initialize enr object with Apis mellifera GAF file
enr$gaf("Apis mellifera")
enr$new()
list.files()
```

As you can see there are now three  files, one Obo file, one Obo data file and
the annotation file.


We will now  create a random  set of genes,  one full set of 100 genes and one
subset  which should  simulate  some  up-regulated  genes. To get at least one
enriched  term  we  will  add two  times  genes  ('Lys-3'  - gene  symbol  and
'LOC113218576' and NCBI identifier) which are  annotated  with the rare
term 'GO:0003796', lysozyme activity.

Here you see the term and its parent nodes:

![](`r goutil$kroki('GO:0003796')`)

```{r}
fullset=sample(unique(enr$data$godata$ID1),88)
## some symbols
# add some non-annotated stupid gene ids
fullset=c(fullset, sprintf("dummy%02i",1:10))
# just select some random genes for enrichment analysis
subset=sample(fullset,8) 
fullset=unique(c(fullset,'Lys-3','LOC113218576'))
subset=unique(c(subset,'Lys-3','LOC113218576'))
length(fullset)
length(subset)
```

Not all genes have  usually  already a GO  annotation  let's check how many of
them have, here we know that we have added 10 dummy genes. To find this out we
should usually convert all gene symbols to NCBI gene  identifiers, for this we
can use the function `enr$symbol2loc`.

```{r}
# How many genes of the subset are in the GAF file
subset
gaf.subset=intersect(enr$symbol2loc(subset),enr$data$godata$ID1)
length(gaf.subset)
gaf.subset
# How many genes of the fullset are in the GAF file
gaf.fullset=intersect(enr$symbol2loc(fullset),enr$data$godata$ID1)
length(gaf.fullset)
```

As you can see genes which are not in the  annotation  file are excluded. This
is as well done internally in the `gaf$enrichment`  function, so do not wonder
why the number of genes used in the  statistics  is not the same as the number
of genes you put into the function. As this elimination of non-annotated genes
is done internally we can put all genes into the analysis.

```{r}
options(width=120)
df=enr$enrichment(fullset,subset)
df$go_name=gsub("(.{20})(.{3,})","\\1 ...",df$go_name)
head(df)
dim(df)
```

Sometimes  we would  like to see not all GO terms, but only some major  terms.
For this the so called  slim-subsets  can be used, here for  instance we could
use the slims for an other insect, *Drosphila melanogaster*.

```{r}
# Now let's only display the drosophila slims
slims=goutil$getSlims(slim="drosophila")
length(slims)
tfslim=rep(FALSE,nrow(df))
tfslim[df$go_id %in% slims]=TRUE
table(tfslim)
### attached slim information
enr=cbind(df,p_slim=tfslim)
### just show slim GO's
enr[enr$p_slim==TRUE,]
```
